---
title: "demo"
output: html_document
---

Reproducible:

"Of 50 of the nation’s largest cities, *Baltimore is one of 34 where police now make homicide arrests less often than in 2014*, according to a Washington Post analysis."

*!Steps Before analysis!*

```{r Libraries}

#install.packages("tidyverse") 
#install.packages("lubridate")
  #uncomment if packages not installed

library(tidyverse)
library(lubridate)

```

```{r Load data}

#data provided by The Washington Post's github
  #link: https://github.com/washingtonpost/data-homicides
homicide_data <- read_csv("homicide-data.csv")

```

```{r Cleaning Dates}
#Need to clean reported_date col first
  #Right now it looks like: <dbl> 20100504 for 2010-05-04

cleaned_homicide_data <- homicide_data |>
  mutate(cleaned_reported_date = ymd(reported_date)) |> #got an error: 2 failed to parse
  mutate(year = year(cleaned_reported_date)) #creating year col for later

head(cleaned_homicide_data) #checking df
```

*1* "Baltimore is one of 34 where police now make homicide arrests less often than in 2014" 

First step is getting the *total number of dispositions by city and year.*

```{r}

all_total_dispositions <- cleaned_homicide_data |>
  group_by(city, year) |>
  summarise(
    total_dispositions = n() #getting totals
  ) |>
  arrange(desc(year)) #arranging from 2017 to smallest date

#all_total_dispositions
  #uncomment to see data frame created

```

*2* Now that we know how many dispositions were made in each city by year we need to find . . . *all homicides closed by arrests by city and year* to the percent of homicide cases are closed by arrests. 

```{r}

all_closed_by_arrest <- cleaned_homicide_data |>
  filter(disposition == "Closed by arrest") |> 
    #filtering by Closed by arrest
  group_by(city, year) |>
  summarise(
    total_closed_by_arrests = n() #getting totals
  )

#all_closed_by_arrest #uncomment to show data frame

```

*3* Now we have two data frames we want to then . . .  *combine the two data frames by city and year* in a >left join< while *calculating the percentage of closed by arrest* for that year.

```{r}

combo_dispo_and_arrest <- left_join(all_total_dispositions, all_closed_by_arrest, by = c("city", "year"), keep = TRUE) |>
  #^ left join by equal city and year values
  #keep = TRUE ensures all data is kept
  mutate(
    year = year.x,
    city = city.x
  ) |> #cleaning column names
  select(-year.x, -year.y, -city.x, -city.y) |> #removing icky names
  mutate(
    total_dispositions = ifelse(is.na(total_dispositions),
                                0, total_dispositions), 
    #making na rows equal to 0 so future calculations do not break            everything
    total_closed_by_arrests = ifelse(is.na(total_closed_by_arrests),
                                     0, total_closed_by_arrests),
    #making NA rows equal to 0 so future cals do not break everything
    percentage = (total_closed_by_arrests / total_dispositions) * 100
    #calculating percent closed by arrest
    ) |>
  arrange(desc(city))

#combo_dispo_and_arrest #uncomment to see data frame

```

*4* Great! We have all of the cases closed by arrest for each major city in the U.S. now we need to . . . *isolating 2014 and 2017 from the rest of the data frames* to compare arrest rates.

I found it was easier to make two separate frames with separate year columns to create a difference column to determine whether a cities 2014 arrest rate increased or decreased compared to 2017.

```{r}
percent_14 <- combo_dispo_and_arrest |>
  filter(year == 2014)

percent_14

percent_17 <- combo_dispo_and_arrest |>
  filter(year == 2017)

percent_17
```

*5* Now we need to join the two data frames by city and create a new column finding the difference between 2017 and 2014.

what we want: 
  difference == 2017_% - 2014_%

what we have: 
  percent_difference = percentage.y [2017] - percentage.x [2014]


```{r}

diff_14_17 <- left_join(percent_14, percent_17, by = "city", keep = TRUE) |>
  mutate(percent_difference = percentage.y - percentage.x,
         #calculating difference
         city = city.x,
         `2014_percent_arrested` = percentage.x,
         `2017_percent_arrested` = percentage.y) |> #cleaning names 
  select(-year.x, -year.y, -city.x, -city.y, -percentage.x, -percentage.y, -total_dispositions.x, -total_dispositions.y, -total_closed_by_arrests.x, -total_closed_by_arrests.y ) #dropping unused/unnecessary names for readability

#diff_14_17 #uncomment to see data frame
```

*6* Now we have the difference between arrest rates! Which cities had a decline and which had an increase in arrest rates?

To find cities with a drop in arrest rates for homicides we need to find which cities have a *negative* difference between 2017 and 2014.

Take this example:

<numbers made up>
Baltimore arrest rate in 2014 = 50
Baltimore arrest rate in 2017 = 27

2017 rate - 2014 rate = rate difference

27% - 50% = -23% difference

Thus, *any negative percent differences indicates a decline in arrest rates between 2014 and 2017 in that city.*

The below count reads back 35 cites instead of the quotes 34 because Durham was counted as having a loss of 63% of arrest rates. Durham can be taken out since there was no rate for city in 2017, therefore the total number of cities that experienced a drop in arrest rates was 34, just like what the WP found.

```{r}

diff_14_17 |>
  filter(
    percent_difference < 0
  ) |>
  nrow()

```